### 2014-12-08 text/Chapter03.tex
Finish proofreading and tweaking (Chapter 3)

In structural coeffect systems, the additional information are associated with individual variables.
This is very often information about how the variables are used, or, in which contexts they are 
used.

In Chapter~\ref{ch:intro}, we introduced the idea using an example that tracks array access patterns.  
Each variable is annotated with a range specifying which elements of the corresponding array 
may be accessed. In this section, we look at a number of examples. We first consider an example
inspired by linear logic. Then we revisit liveness and data-flow, for which the structural system 
provides a more precise analysis. Finally, we look at a number of other practical uses including 
security, tainting and provenance tracking.

@@

The flat system for liveness analysis presented in Section~\ref{sec:applications-flat-live} is 
interesting from a theoretical perspective, but it is not practically useful. In this section, we
revisit the problem and define a structural system that tracks liveness per-variable.

@@

\begin{itemize}
\item The fact that variables are now a vector means that we cannot freely re-order them. This
  guarantees that $\coctx{x\!:\!\tau, y\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$
  can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$.
  We need to define the type system in a way that is similar to sub-structural systems 
  (discussed in Section~\ref{sec:path-logic}) which provide explicit rules for manipulating
  the context.

\item We choose to attach a vector of annotations to a vector of variables, rather than attaching
  individual annotations to individual variables. This lets us unify and combine flat and 
  structural systems as discussed in Chapter~\ref{ch:unified}, but the alternative is briefly
  explored in Section~\ref{sec:conclusions-meta}.
\end{itemize}

\paragraph{Type system.}
The structural system for liveness uses the same two-point lattice of annotations
$\mathcal{L}=\{ \ident{L}, \ident{D} \}$ that was used by the flat system. We also use the
$\sqcup, \sqcap$ and $\sqsubseteq$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.

The rules of the system are split into two groups. Figure~\ref{fig:applications-struct-live} (a) shows 
the standard syntax-driven rules plus sub-coeffecting. In (\emph{var}), the context contains just the 
single accessed variable, which is annotated as live. Other variables can be introduced using weakening. 
A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
sub-coeffecting rule (\emph{sub}) uses a pointwise extension of the $\sqsubseteq$ relation over two
vectors as defined in Section~\ref{sec:applications-strucutre-vec}.

@@

The context $\Gamma_1$ is used to evaluate $e_1$ and is thus annotated with $\aclrd{\mathbf{r}}$. 
The annotation for $\Gamma_2$ is more interesting. It is a result of sequential composition of two
semantic functions -- the first one takes the (multi-variable) context $\Gamma_2$ and evaluates 
$e_2$; the second takes the result of type $\tau_1$ and passes it to the function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$.
The composition is defined as follows:
%
\begin{equation*}
g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\aclrd{\mathbf{s}}} \sigma
\qquad
f : \sigma \xrightarrow{\cclrd{t}} \tau
\qquad
f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \sqcup \aclrd{\mathbf{s}}} \tau
\end{equation*}
%
This definition is only for illustration and is revised in Chapter~\ref{ch:structural}. The function
$g$ takes a product of multiple variables (and is annotated with a vector). The function $f$ takes
just a single value and is annotated with the scalar. As in the flat system, sequential composition
is modelled using $\sqcup$ -- but here, we use a scalar-vector extension of the operation. Finally,
the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$.

@@

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c:\tau \in \Delta}
  {\coctx{()}{\alift{}} \vdash c : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\vspace{0.5em}

@@

@@

\paragraph{Semantics.}
When defining the semantics of flat liveness calculus, we used an indexed form of the option type
$1 + \tau$ (which is $1$ for dead contexts and $\tau$ for live contexts). In the semantics of 
expressions, the type wrapped the entire context, \ie~$1+(\tau_1 \times \ldots \times \tau_n)$.
In the structural version, the semantics wraps individual elements of the free-variable context
pair: $(1+\tau_1) \times \ldots \times (1+ \tau_n)$. For each variable, the type is indexed by 
the corresponding annotation. More formally:
%
\begin{equation*}
\begin{array}{r}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
  ~:~ (\tau'_1 \times \ldots \times \tau'_n) \rightarrow \tau\\[0.5em]
\textnormal{where}~\tau'_i = \begin{cases}
\tau_i & (\cclrd{r_i} = \cclrd{\ident{L}})\\
1      & (\cclrd{r_i} = \cclrd{\ident{D}})
\end{cases}
\end{array}
\end{equation*}
%
Note that the product of the free variables is not an ordinary tuple of our language, but a special
construction. This follows from the asymmetry of $\lambda$-calculus, as discussed in 
Section~\ref{sec:applications-strucutre-vec}. Functions take just a single input and so they are
interpreted in the same way as in flat calculus:
%
\begin{equation*}
\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2 \qquad\qquad
\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
\end{equation*}
%
The rules that define the semantics are shown in Figure~\ref{fig:applications-struct-livesem}.
To make the definition simpler, we are somewhat vague when working with products. We write 
variables of product type such as $\mathbf{v}$ in bold-face and individual values like $x$ in 
normal face. We freely re-associate products and so $(\mathbf{v}, x)$ should not be seen as a
nested product, but simply a product with a number of variables represented as another product
$\mathbf{v}$ with one more variable $x$ at the end. We shall be more precise in Chapter~\ref{ch:structural}.

In application, we again need to ``implement'' dead code elimination. When the input parameter
of the function $g$ is live (\emph{app-1}), we first evaluate $e_2$ and then pass the result to
$g$. When the parameter is dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values
in $\mathbf{v_2}$ can be dead, \ie~$()$.

\paragraph{Summary.}
The structural liveness calculus is a typical example of a system that tracks per-variable 
annotations. In a number of ways, the system is simpler than the flat coeffect calculi. In
lambda abstraction, we simply annotate function with the annotation of a matching variable 
(this rule is the same for all upcoming systems). In application, the \emph{pointwise} composition
is no longer needed, because the sub-expressions use separate contexts. On the other hand, 
we had to add weakening, contraction and exchange rules to let us manipulate the contexts.

The semantics of weakening demonstrates an important point about coeffects that may be quite
confusing. When we read the \emph{typing rule} from top to bottom, weakening adds a variable
to the context. When we read the \emph{semantic rule}, weakening drops a variable value from the 
context! This duality is caused by the fact that coeffects talk about context -- they describe
how to build the context required by the sub-expressions and so the semantics implements 
transformation from the context in the (typing) conclusion to the (typing) assumption.

The structural systems discussed in the upcoming sections are remarkably similar to the one
shown here. We discuss two more examples in details to explore the design space, but we shall
omit details that are the shared with the system in this section.

@@

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\vspace{0.5em}

@@

\paragraph{Type system.}
The type system in Figure~\ref{fig:applications-struct-bll} annotates contexts with vectors of integers.
The rules have the same structure as those of the system for liveness analysis. The only difference is
how annotations are combined -- here, we use integer multiplication ($\ast$) and addition ($+$).

As previously, application (\emph{app}) and let binding (\emph{let}) combine two separate contexts.
The second part applies a function that uses its parameter $\cclrd{t}$-times to an argument that uses
variables in $\Gamma_2$ at most $\aclrd{\mathbf{s}}$-times (here, $\aclrd{\mathbf{s}}$ is a vector of
integers with an annotations for each variable in $\Gamma_2$). The sequential composition (modelling
call-by-name) multiplies the uses, meaning that the total number of uses is $(\cclrd{t} \ast \aclrd{\mathbf{s}})$
(where $\ast$ is a multiplication of a vector by a scalar). This models the fact that for each use of
the function parameter, we replicate the variable uses in $e_2$.

Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
the number of uses using the pointwise
$\leq$ relation. We can view any variable as being used a greater
number of times than it actually is.

@@

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian closed 
category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

@@

\paragraph{Bounded linear logic.}
The system presented in this section differs from bounded linear logic (BLL) \cite{logic-bounded}.
Using the terminology from Section~\ref{sec:path-sem-contextdep}, our system is written in the
\emph{language semantics} style, while BLL is written in the \emph{meta-language} style.

This means that the terms and types of our system are the terms and types of ordinary 
$\lambda$-calculus, with the only difference that functions carry coeffect annotations.
In BLL, the language of types is extended with a type constructor $!_k A$ (where $A$ is
a proposition, corresponding to a type $\tau$ in our system). The type denotes a value
$A$ that can be used at most $k$ times.

As a result, BLL does not need to attach additional annotation to the variable context 
as a whole. The requirements are attached to individual variables and so our context
$\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$ corresponds
to a BLL assumption $!_{k_1} A_1, ..., !_{k_n} A_n$.

Using the formulation of bounded logic (and omitting the terms), the weakening and
contraction rules are written as follows:
%
\[
\tyrule{weak}
  {\Gamma \vdash B}
  {\Gamma, !_0 A \vdash B}
\quad
\tyrule{contr}
  {\Gamma, !_n A, !_m A \vdash B}
  {\Gamma, !_{n+m} A \vdash B}
\]
%
The system captures the same idea as the structural coeffect system presented above.
Variable access in bounded linear logic is simply an operation that produces a value
$!_n A$ and so the system further introduces \emph{dereliction} rule which lets us 
treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
semantics} and \emph{meta-language} in Section~\ref{sec:conclusions-meta}.

@@

The brief overview of bounded linear logic shows an alternative approach to tracking properties
related to individual variables -- we could attach annotations to the variables themselves 
rather than attaching a \emph{vector} of annotations to the entire context. Our approach has
two benefits -- it lets you unify flat and structural systems (Chapter~\ref{ch:unified}) and 
it also makes it possible to build composed systems that mix both flat and structural properties.

@@

When discussing data-flow languages in the previous section, we said that the context provides 
past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as 
a \emph{flat} property, which gives us a system that keeps the same number of past values for
all variables. However, data-flow can also be adapted to a structural system which keeps the number 
of required past values individually for each variable. Consider the 
following example:
%
\begin{equation*}
\kvd{let}~\ident{offsetZip} = \ident{left} + \kvd{prev}~\ident{right}
\end{equation*}
%
The value \ident{offsetZip} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
requires $1$ past values of both streams in the variable context.

Turning a flat data-flow system to a structural data-flow system is a change similar to the one
between flat ans structural liveness. In case of liveness analysis, we included the flat system
only as an illustration (it is not practically useful).

For data-flow, the flat system is less precise, but still practically useful (simplicity may outweigh 
precision). As discussed in Section~X, the structural system is necessary when we allow arbitrary 
recursion and cannot (easily) determine the number of required values statically.

@@

\paragraph{Semantics.}
To define the semantics of our structural data-flow language, we can use the same approach as when
adapting flat liveness to structural liveness. Rather than wrapping the whole context in some wrapper
(list or option type), we now wrap individual components of the product representing the variables
in the context.

@@

\paragraph{Summary.}
Two of the structural examples shown so far (liveness and data-flow) extend an earlier flat 
version of a similar system. We discuss this relation in general later. However, a flat system 
can generally be turned into a structural one -- although this only gives a useful system when
the flat version captures properties related to variables.

@@

\subsection{Security, tainting and provenance}
Tainting is a mechanism where variables coming from potentially untrusted sources are marked
(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
can cause security issues or other problems. Tainting can be done dynamically as a runtime mark
(\eg~in the Perl language) or statically using a type system. Tainting can be viewed as a special
case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
values are annotated with more detailed information about their source.

>>>>

In structural coeffect systems, the additional information is associated with individual variables.
This is very often information about how the variables are used, or, in which contexts they are used. 
In Chapter~\ref{ch:intro}, we introduced the idea using an example that tracks array access patterns.  
Each variable is annotated with a range specifying which elements of the corresponding array 
may be accessed.

In this section, we look at three examples in detail -- we revisit liveness and show a practically
useful structural version of the system; we consider an example inspired by linear logic; finally,
we revisit data-flow to get a more precise analysis. Although quite different, the common pattern
among these three examples is somewhat easier to see, because they all track information about 
variable usage. We finish the section with a brief outline of several other applications.

@@

The flat system for liveness analysis presented in Section~\ref{sec:applications-flat-live} is 
interesting from a theoretical perspective, but it is not practically useful. Here, we
revisit the problem and define a structural system that tracks liveness per-variable.

@@

\begin{itemize}
\item The fact that variables are now a vector means that we cannot freely re-order them. This
  guarantees that $\coctx{x\!:\!\tau, y\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$
  can not be confused with $\coctx{y\!:\!\tau, x\!:\!\tau}{\alift{\cclrd{\ident{L}}, \cclrd{\ident{D}}}}$.
  We need to define the type system in a way that is similar to sub-structural systems 
  (discussed in Section~\ref{sec:path-logic}) and add explicit rules for manipulating
  the context.

\item We choose to attach a vector of annotations to a vector of variables, rather than attaching
  individual annotations to individual variables. This lets us unify and combine flat and 
  structural systems as discussed in Section~\ref{sec:unified-unified}, but the alternative is briefly
  explored in Section~\ref{sec:unified-meta}.
\end{itemize}

\paragraph{Type system.}
The structural system for liveness uses the same two-point lattice of annotations
$\mathcal{L}=\{ \cclrd{\ident{L}}, \cclrd{\ident{D}} \}$ that was used by the flat system. We also use the
$\sqcup, \sqcap$ and $\sqsubseteq$ operators that are defined in Figure~\ref{fig:applications-flat-livealg}.

The rules of the system are split into two groups. Figure~\ref{fig:applications-struct-live} (a) shows 
the standard syntax-driven rules plus sub-coeffecting. In (\emph{var}), the context contains just the 
single accessed variable, which is annotated as live. Unused variables can be introduced using weakening. 
A constant (\emph{const}) is accessed in an empty context, which also carries no annotations. The
sub-coeffecting rule (\emph{sub}) uses a pointwise extension of the $\sqsubseteq$ relation over two
vectors as defined in Section~\ref{sec:applications-strucutre-vec}.

@@

The context $\Gamma_1$ is used to evaluate $e_1$ and is thus annotated with $\aclrd{\mathbf{r}}$. 
The annotation for $\Gamma_2$ is more interesting. It is a result of sequential composition of two
semantic functions -- the first one takes the (multi-variable) context $\Gamma_2$ and evaluates 
$e_2$; the second takes the result of type $\tau_1$ and passes it to the function $\tau_1 \xrightarrow{\cclrd{t}} \tau_2$.
The composition is defined as follows:
%
\begin{equation*}
g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\aclrd{\mathbf{s}}} \sigma
\qquad
f : \sigma \xrightarrow{\cclrd{t}} \tau
\qquad
f \circ g : \tau_1 \times \ldots \times \tau_n \xrightarrow{\cclrd{t} \,\sqcup\, \aclrd{\mathbf{s}}} \tau
\end{equation*}
%
This definition is only for illustration and is revised in Chapter~\ref{ch:structural}. The function
$g$ takes a product of multiple variables (and is annotated with a vector). The function $f$ takes
just a single value and is annotated with the scalar. As in the flat system, sequential composition
is modelled using $\sqcup$ -- but here, we use a scalar-vector extension of the operation. Finally,
the (\emph{let}) rule follows similar reasoning (and also corresponds to the typing of $(\lambda x.e_2)~e_1$).

@@

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules along with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{\ident L}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c:\tau \in \Delta}
  {\coctx{()}{\alift{}} \vdash c : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \sqcup \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \sqsubseteq \aclrd{\textbf{r'}}
\end{equation*}
\vspace{0.5em}

@@

\paragraph{Semantics.}
When defining the semantics of flat liveness calculus, we used an indexed form of the option type
$1 + \tau$ (which is $1$ for dead contexts and $\tau$ for live contexts). In the semantics of 
expressions, the type constructor was applied to the entire context, \ie~$1+(\tau_1 \times \ldots \times \tau_n)$.
In the structural version, the semantics applies the option type constructor to individual elements 
of the free-variable context pair: $(1+\tau_1) \times \ldots \times (1+ \tau_n)$. For each variable, 
the type is indexed by the corresponding annotation:
%
\begin{equation*}
\begin{array}{r}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \alift{\cclrd{r_1}, \ldots, \cclrd{r_n}} } \vdash e : \tau} 
  ~:~ (\tau'_1 \times \ldots \times \tau'_n) \rightarrow \tau\\[0.5em]
\textnormal{where}~\tau'_i = \begin{cases}
\tau_i & (\cclrd{r_i} = \cclrd{\ident{L}})\\
1      & (\cclrd{r_i} = \cclrd{\ident{D}})
\end{cases}
\end{array}
\end{equation*}
%
Note that the product of the free variables is not an ordinary tuple of our language, but a special
construction (we return to this topic in Section~\ref{sec:struct-semantics}). This follows from the 
asymmetry of $\lambda$-calculus, as discussed in Section~\ref{sec:applications-strucutre-vec}. Functions 
take just a single input and so they are interpreted in the same way as in flat calculus:
%
\begin{equation*}
\sem{\tau_1 \xrightarrow{\aclrd{\ident{L}}} \tau_2} = \tau_1 \rightarrow \tau_2 \qquad\qquad
\sem{\tau_1 \xrightarrow{\aclrd{\ident{D}}} \tau_2} = 1 \rightarrow \tau_2
\end{equation*}
%
The rules that define the semantics are shown in Figure~\ref{fig:applications-struct-livesem}.
To make the definition simpler, we are somewhat vague when working with products. We write 
variables of product type such as $\mathbf{v}$ in bold-face and individual values like $x$ in 
normal face. We freely re-associate products and so $(\mathbf{v}, x)$ should not be seen as a
nested product, but simply as a product containing all variables from the product $\mathbf{v}$ 
together with one additional variable $x$ at the end. We shall be more precise in 
Chapter~\ref{ch:structural}.

As dictated by the semantics, the application again needs to ``implement'' dead code elimination
(otherwise the type system would be unsound). When the input parameter of the function $g$ is live 
(\emph{app-1}), we first evaluate $e_2$ and then pass the result to $g$. When the parameter is 
dead (\emph{app-2}), we do not need to evaluate $e_2$ and so all values in $\mathbf{v_2}$ can be 
dead, \ie~$()$.

@@

\paragraph{Summary.}
The structural liveness calculus is a typical example of a system that tracks per-variable 
annotations. In a number of ways, the system is simpler than the flat coeffect calculi. In
lambda abstraction, we simply annotate function with the annotation of a matching variable 
(this rule is the same for all upcoming systems). In application, the \emph{pointwise} composition
is no longer needed, because the sub-expressions use separate contexts. On the other hand, 
we had to add weakening, contraction and exchange rules to let us manipulate contexts.

The semantics of weakening demonstrates an important point about coeffects that may be quite
confusing. When we read the \emph{typing rule} from top to bottom, weakening adds a variable
to the context. When we read the \emph{semantic rule}, weakening drops a variable value from the 
context! This duality is caused by the fact that coeffects talk about context -- they describe
how to build the context required by the sub-expressions and so the semantics implements 
transformation from the context in the (typing) conclusion to the (typing) assumption. How
should coeffects be understood, in general, is discussed further in Section~\ref{sec:flat-calculus-undestanding}.

The structural systems discussed in the upcoming sections are remarkably similar to the one
shown here. We discuss two more examples to explore the design space, but we shall
omit details that are the shared with the system in this section.

@@

\paragraph{Type system.}
The type system in Figure~\ref{fig:applications-struct-bll} annotates contexts with vectors of integers.
The rules have the same structure as those of the system for liveness analysis. The only difference is
how annotations are combined. Here, we use integer multiplication ($\ast$) for sequential composition
and addition ($+$) for point-wise composition.

As previously, application (\emph{app}) and let binding (\emph{let}) combine two separate contexts.
The second part applies a function that uses its parameter $\cclrd{t}$-times to an argument that uses
variables in $\Gamma_2$ at most $\aclrd{\mathbf{s}}$-times (here, $\aclrd{\mathbf{s}}$ is a vector of
integers with an annotations for each variable in $\Gamma_2$). The sequential composition (modelling
call-by-name) multiplies the uses, meaning that the total number of uses is $(\cclrd{t} \ast \aclrd{\mathbf{s}})$
(where $\ast$ is a point-wise multiplication of a vector by a scalar). This models the fact that 
for each use of the function parameter, we replicate the variable uses in $e_2$.

Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates the number of accesses using 
the pointwise $\leq$ relation. We can view any variable as being used a greater
number of times than it actually is.

\begin{figure}[t]
{\small a.) Ordinary, syntax-driven rules along with sub-coeffecting}
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau} 
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2} 
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad 
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\vspace{0.5em}

@@

@@

\paragraph{Semantics.}
In the previous examples, we defined the semantics -- somewhat informally -- using a simple 
$\lambda$-calculus language to encode the model. More formally, this could be a Cartesian Closed 
Category. In that model, we can reuse variables arbitrarily and so it is not
a good fit for modelling bounded reuse. Girard et al. \cite{logic-bounded} model their bounded
linear logic in an (ordinary) linear logic where variables can be used at most once.

@@

\paragraph{Bounded linear logic.}
The system presented in this section is based on the idea of bounded linear logic (BLL)
\cite{logic-bounded}, but it is adapted to follow the structure of other coeffect systems 
discussed in this chapter. This elucidates the connection between BLL and coeffects.

The big difference, using the terminology from Section~\ref{sec:path-sem-contextdep}, is 
that our system is written in \emph{language semantics} style, while BLL is written 
in \emph{meta-language} style. We briefly consider the original BLL formulation.

The terms and types of our system are the terms and types of an ordinary $\lambda$-calculus, 
with the only difference that functions carry coeffect annotations. In BLL, the language of 
types is extended with a type constructor $!_k A$ (where $A$ is a proposition, corresponding 
to a type $\tau$ in our system). The type denotes a value $A$ that can be used at most $k$ times.

As a result, BLL does not need to attach additional annotation to the variable context 
as a whole. The requirements are attached to individual variables and so our context
$\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$ corresponds
to a BLL assumption $!_{k_1} A_1, ..., !_{k_n} A_n$. Using the formulation of bounded logic 
(and omitting the terms), the weakening and contraction rules are written as follows:
%
\[
\tyrule{weak}
  {\Gamma \vdash B}
  {\Gamma, !_0 A \vdash B}
\quad
\tyrule{contr}
  {\Gamma, !_n A, !_m A \vdash B}
  {\Gamma, !_{n+m} A \vdash B}
\]
%
The system captures the same idea as the structural coeffect system presented above.
Variable access in bounded linear logic is simply an operation that produces a value
$!_n A$ and so the system further introduces \emph{dereliction} rule which lets us 
treat $!_1 A$ as a value $A$. We further explore difference between \emph{language
semantics} and \emph{meta-language} in Section~\ref{sec:unified-meta}.

@@

The brief overview of bounded linear logic shows an alternative approach to tracking properties
related to individual variables -- we could attach annotations to the variables themselves 
rather than attaching a \emph{vector} of annotations to the entire context. The main benefit
of our approach is that it lets us unify flat and structural systems (Chapter~\ref{ch:unified}).

@@

When discussing data-flow languages in an earlier section, we said that the context provides 
past values of variables. In Section~\ref{sec:applications-flat-dataflow}, we tracked this as 
a \emph{flat} property, which gives us a system that keeps the same number of past values for
all variables. However, data-flow can also be adapted to a structural system which keeps the number 
of required past values individually for each variable. Consider the 
following example:

\begin{equation*}
\kvd{let}~\ident{offsetAdd} = \ident{left} + \kvd{prev}~\ident{right}
\end{equation*}
%
The value \ident{offsetAdd} adds values of \ident{left} with previous values of \ident{right}.
To evaluate a current value of the stream, we need the current value of \ident{left} and one past
value of \ident{right}. Flat system is not able to capture this level-of-detail and simply
requires $1$ past values of both streams in the variable context.

Turning a flat data-flow system to a structural data-flow system is a change similar to the one
between flat ans structural liveness. In case of liveness analysis, we included the flat system
only as an illustration (it is not practically useful). For data-flow, the flat system is less 
precise, but still practically useful (simplicity may outweigh precision).

@@

\paragraph{Semantics.}
To define the semantics of our structural data-flow language, we can use the same approach as when
adapting flat liveness to structural liveness. Rather than wrapping the whole context in a type 
constructor (list or option), we now wrap the individual components of the product representing 
the variables in the context.

@@

\paragraph{Summary.}
Two of the structural examples shown so far (liveness and data-flow) extend an earlier flat 
version of a similar system. We discuss this relation in general later. However, a flat system 
can generally be turned into a structural one -- although this only gives a useful system when
the flat version captures statically scoped properties, \ie~related to variables.

@@

\subsection{Security, tainting and provenance}
Tainting is a mechanism where variables coming from potentially untrusted sources are marked
(\emph{tainted}) and the use of such variables is disallowed in contexts where untrusted input
can cause security issues or other problems. Tainting can be done dynamically using a runtime mark
(\eg~in the Perl language) or using a static type system. Tainting can be viewed as a special
case of \emph{provenance tracking}, known from database systems \cite{app-provenance-db}, where
values are annotated with more detailed information about their source.
